---
title: "Observation weights and missing data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Observation weights and missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rwa)
```

## Weighted point estimates

Use `weight` to name a numeric column of observation weights. Non-missing
weights must be finite and strictly positive; zero weights are not supported.
The outcome and predictors must be numeric. Weighted RWA decomposes the
R-squared from the weighted correlation matrix, matching weighted least-squares
regression with an intercept on the same retained rows.

```{r weighted-estimate}
d <- mtcars
d$observation_weight <- rep(c(1, 2, 5, 3), 8)
fit <- rwa(d, "mpg", c("hp", "wt", "disp"), method = "multiple",
           weight = "observation_weight", sort = FALSE)
fit$result
c(rwa = fit$rsquare,
  weighted_lm = summary(lm(mpg ~ hp + wt + disp, data = d,
                           weights = observation_weight))$r.squared)
```

Multiplying every weight by the same positive constant leaves relative weights,
R-squared, and `n_effective` unchanged. Integer weights also give the same point
estimates as repeating each row that many times. These point-estimate
equivalences do not imply equivalent bootstrap sampling designs.

```{r weight-scale}
d$scaled_weight <- 100 * d$observation_weight
scaled <- rwa_multiregress(d, "mpg", c("hp", "wt", "disp"), weight = "scaled_weight")
all.equal(fit$result, scaled$result)
c(original_effective = fit$n_effective, scaled_effective = scaled$n_effective)
```

## Missing-data contract

Rows with missing outcomes are **always removed first**, even with
`use = "all.obs"` or `"everything"`. For unweighted analysis, `use` then applies
to correlations among the retained outcome and predictors:

| `use` | Unweighted behavior after removing missing outcomes |
|---|---|
| `"pairwise.complete.obs"` (default) | Use available pairs for each correlation. An incompatible joint matrix causes an error. |
| `"complete.obs"` | Use rows complete on all analysis variables. |
| `"na.or.complete"` | Use complete rows when available. `cor()` would return NAs when none exist; RWA instead reports insufficient data. |
| `"all.obs"` | Error if remaining predictors contain missing values. |
| `"everything"` | Missing values propagate to correlations; a non-finite matrix causes an informative RWA error. |

With weights, the default is effectively **complete-case analysis**, not
weighted pairwise correlation. After removing missing outcomes:

1. Missing weights cause an error for `"all.obs"` and are removed for every
   other mode.
2. Rows with missing predictors are removed for **every weighted mode**,
   including `"all.obs"` and `"everything"`.

These exceptions preserve the package's existing preprocessing. `all.obs`
does not mean that every value in the original input must be observed. Weights
are validated for type, finiteness, and positivity before row filtering.

```{r missing-data}
d$mpg[1] <- NA
d$observation_weight[2] <- NA
d$hp[3] <- NA
missing_fit <- rwa(d, "mpg", c("hp", "wt", "disp"), method = "multiple",
                   weight = "observation_weight")
retained <- complete.cases(d[c("mpg", "hp", "wt", "disp", "observation_weight")])
c(n = missing_fit$n, retained_rows = sum(retained),
  n_weighted = missing_fit$n_weighted,
  retained_weight_sum = sum(d$observation_weight[retained]))
```

### Counts and effective sample size

- `n` retains its existing meaning: the complete-case count for the selected
  variables (and weight, if provided). Unweighted pairwise correlations may
  use more observations than this conservative count.
- `n_weighted` is included only for weighted results and sums **original**
  weights after outcome, missing-weight, and predictor-completeness filters.
  It represents population size only when weights are calibrated to that scale
  and the retained rows represent the intended population scope. Arbitrarily
  normalized weights do not supply a population count.
- `n_effective` is included only for weighted results and is Kish's
  unequal-weighting effective sample size: `(sum(w)^2) / sum(w^2)`. The
  calculation first scales weights by their maximum to avoid overflow or
  underflow when squaring them. It ignores clustering, stratification, and
  relationships between weights and outcomes. It is **not** model degrees of
  freedom, exact RWA precision, or a complete survey design effect.

Both exported paths, `rwa()` and `rwa_multiregress()`, return these weighted-only
diagnostics. Unweighted result names and structure are unchanged.

## Bootstrap scope and failure policy

`bootstrap = TRUE` performs independent, identically distributed (iid)
**individual-row resampling with replacement**. Each sampled row carries its
original weight, and sampling is not proportional to weights. Missing outcomes
are removed before resampling, but missing-weight and predictor filters apply
**within each sample**. Consequently, the sampling frame can be larger than the
point estimate's complete-case data and individual samples can retain different
numbers of rows.

A weight column does not encode clusters, strata, selection dependence, or
replicate-weight designs. This interface is observation-weighted RWA with an
iid-row bootstrap, not general complex-survey variance estimation. The
appropriateness of ordinary row resampling depends on the sampling design;
use design-aware methods when the iid-row assumptions are unsuitable.

```{r bootstrap, eval=FALSE}
set.seed(42)
boot_fit <- rwa(d, "mpg", c("hp", "wt", "disp"), method = "multiple",
                weight = "observation_weight", bootstrap = TRUE,
                n_bootstrap = 1000, comprehensive = TRUE)
boot_fit$bootstrap$ci_results$random_comparison
```

Comprehensive analysis compares each requested predictor to an added random
predictor, even without a focal predictor. Supplying, for example,
`focal = "wt"` additionally compares the other predictors to `wt`. Interval
labels follow the requested predictor order, independently of result sorting.

Every sample must estimate the same requested model. A constant column,
singular predictor matrix, or invalid joint matrix causes a clear error and
stops the bootstrap. Samples are **not skipped or retried**, predictors are not
dropped, and statistics are not recycled. A rare binary predictor that loses
its minority category in a sample is one such failure. Inspect predictor
variation, sample size, and missingness rather than treating incomplete
bootstrap output as valid intervals.

## Matrix diagnostics

The full joint outcome/predictor correlation matrix must be finite, have the
expected variable identities and dimensions, and be positive semidefinite
within tolerance. Its smallest eigenvalue must be at least minus
`sqrt(.Machine$double.eps) * max(1, max(abs(eigenvalues)))`. The predictor block
must have strictly positive computed eigenvalues and its transformation must
be solvable. No additional conditioning cutoff is imposed on previously
estimable models. Highly correlated predictors can still produce sensitive
estimates, so inspect their stability rather than treating successful numerical
execution as evidence of reliable individual weights.

Calculated R-squared is also checked: values above
`1 + sqrt(.Machine$double.eps)` produce an error, not a clipped result.
This separate check is needed because nearly collinear predictors can amplify
a small joint-matrix error into a materially impossible fit.

A singular **joint** matrix is valid for an exact fit when the **predictor**
block remains invertible. A materially indefinite joint matrix is not valid,
even if its predictor block is invertible: pairwise correlations computed on
different row subsets can otherwise produce an impossible R-squared above one.
Consider `use = "complete.obs"` and inspect missingness if this occurs; changing
the retained sample changes the analysis and is not an automatic repair.

No eigenvalue clipping, matrix repair, or automatic change to complete cases is
performed. Constant variables, insufficient observations, and zero R-squared
(for which rescaled relative weights are undefined) also produce informative
errors.
